---
title: "DAM AT1 Part A Linear Regression, by Alex Brooks"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#note to self: correct working directory is setwd("/Users/alex/Projects/DAMAssignment1")
#load libraries
library(readr)
library(ggplot2)
library(dplyr)
```

### AT1 Spring 2018: Part A
### Data Cleaning and Exploratory Data Analysis 
```{r}
#import dataset. Create a function as we may want to do this again later for a clean import.

read_and_clean_data <- function(datasource, hasHeader=T) {
  dataSet = read.csv(datasource, header = hasHeader)
  #check data structures match the supplied data dictionary

  #Clean data columns to match data dictionary
  dataSet$date = as.Date(dataSet$date, format = "%d/%m/%y")
  dataSet$customer_id = as.character(dataSet$customer_id, format = "")
  
  return(dataSet)
}

transactions <- read_and_clean_data("transactions.csv", T)

```

## 1. a. Cleaning the data

The values in the date column should be in Date format, not Factor.
Customer_id needs to be String or Character, not Factor.  

```{r}
#View the data and its structure structure to confirm
str(transactions)
View(transactions)
```

## EDA: histograms of industry & location variables

```{r}
par(mfrow=c(1,2))
hist(transactions$industry, main = "Trans by Industry", xlab="Industry", ylab="Number of transactions", xlim = c(0,10), ylim=c(0,50000), las=0)
hist(transactions$location, main = "Trans by Location", xlab="Location", ylab="Number of transactions", xlim = c(0,10), ylim=c(0,50000), las=0)
```


```{r}
#let's visualise a scatterplot showing industry against monthly amount over time
ggplot(transactions, aes(x=date, y=monthly_amount, color=industry)) + 
  geom_point() + geom_smooth(se = FALSE, color = "red") + labs(title="Industry and monthly amount over time", x="Date", y="Monthly Amount")

```
```{r}
#let's visualise a scatterplot showing location variables against monthly amount over time
ggplot(transactions, aes(x=date, y=monthly_amount, color=location)) + 
  geom_point() + geom_smooth(se = FALSE, color = "red") + labs(title="Location and monthly amount over time", x="Date", y="Monthly Amount")
```
## What about exploring detail about higher than average sales values?
It makes sense to explore where and why the monthly amounts are higher than average, using the 'mean' as the indicator of average.
```{r}
#Mutate the dataset to create a new column called 'monthly_amount_above_mean'  
transactions <- transactions %>%
  # creating a new variable to classify transaction column
  mutate(monthly_amount_above_mean = ifelse(monthly_amount > mean(monthly_amount), TRUE, FALSE))
```

```{r}
# plot a bar chart by monthly amounts by date  
qplot(x = date, fill = monthly_amount_above_mean, data = transactions, geom = "bar", main = "Number of transactions by month, showing proportion above the mean")
```

This reveals most monthly amounts are below the mean and the monthly amounts grow steadily rather than steeply. Let's now look at it by industry and location, too

```{r}
# plot a bar chart by monthly amounts by location
qplot(x = location, fill = monthly_amount_above_mean, data = transactions, geom = "bar", main = "Number of transactions by location, showing proportion above the mean")
```

## So what about by industry?

```{r}
# plot a bar chart by monthly amounts by industry
qplot(x = industry, fill = monthly_amount_above_mean, data = transactions, geom = "bar", main = "Number of transactions by industry, showing proportion above the mean")
```
## 1. b. Description of 2 insights from EDA
### Some insights for the sales manager
Location 1 and 2 have the highest volumes of monthly_amounts, as well as the most above-average monthly amounts.

Industry 6 has the least amount of sales transactions but despite its low sales volume, it has above average sales amounts. 

Industries 3, 5 and 9 seem to have a greater proportion of their transactions above the mean average monthly amount

### Basic model fitting
## 2. a. Creating the model
The goal wil be to build a model that can predict future monthly_amounts. We will need to add a new variable called "TIME", which gives values of 1 to 47 to each month and year in the dataset. For example: January, 2013 becomes 1, February 2013 becomes 2 until November 2016 becomes 47.

Time = 1,2,3,4 ... 46, 47

We will then do a rough 70:30 test:train data split, with T = 1,2 ..31, 32 in the training set and 33,34 ...46,47 in the test set.

We potentially need to factor in the yearly growth trend to remove it from our test/train data.

## 2. a. i) Aggregate the data  

###ANTHONY TO ADD TIME VARIABLE HERE


```{r}
# Aggregate the data, grouping by date, industry and location, and calculating the mean monthly_amount

#we may want to do this again, so write a function
aggregate_transactions <- function(datasource, hasHeaders = T) {
  
  #Use the function we wrote earlier
  dataSet = read_and_clean_data(datasource, hasHeaders)
    
  # Aggregate the data, grouping by date, industry and location, and calculating the mean monthly_amount
  output = dataSet %>%
    group_by(date, industry, location) %>%
    summarize(monthly_amount = mean(monthly_amount, na.rm = TRUE))
  
  # Let's also create a column for the month number and another one for year number
  output = output %>%
    mutate(month_number = format(as.Date(date), "%m")) %>%
    mutate(year_number = format(as.Date(date), "%Y"))
  
  #Make sure the new columns are of the correct type
  output$month_number = as.integer(output$month_number)
  output$year_number = as.integer(output$year_number)
  
  transform(output, month_number = as.integer(month_number), year_number = as.integer(year_number))
  
  return(output)

}

aggregated_transactions <- aggregate_transactions("transactions.csv", T)

View(aggregated_transactions)
```
###ANT - we will want TIME to appear in that dataset above
## 2. a. ii. Create a line plot of the variable monthly_amount for industry = 1 and location = 1

```{r}
#Create a line plot of the variable monthly_amount for industry = 1 and location = 1
ggplot(data=filter(aggregated_transactions, industry == 1 & location == 1), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Industry 1 and Location 1", x="Date", y="Mean Monthly Amount")
```

## What about line plots of other industry and location codes?
```{r}
#Create a line plot of the variable monthly_amount for industry = 2 and location = 2
ggplot(data=filter(aggregated_transactions, industry == 2 & location == 2), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Industry 2 and Location 2", x="Date", y="Mean Monthly Amount")
```
The line plot reveals high monthly seasonality, with November and December mean monthly sales mounts plummeting compared to January and February.

## What about line plots of other industry and location codes?
```{r}
# Create a line plot of the variable monthly_amount for industry = 3 and location = 3
ggplot(data=filter(aggregated_transactions, industry == 3 & location == 3), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Industry 3 and Location 3", x="Date", y="Mean Monthly Amount")
```

# Create a line plot of the variable monthly_amount for industry = 4 and location = 4
```{r}
ggplot(data=filter(aggregated_transactions, industry == 4 & location == 4), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Industry 4 and Location 4", x="Date", y="Mean Monthly Amount")
```
```{r}
ggplot(data=filter(aggregated_transactions, industry == 5 & location == 5), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Industry 5 and Location 5", x="Date", y="Mean Monthly Amount")
```
```{r}
ggplot(data=filter(aggregated_transactions, location == 6), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Location 6 ONLY", x="Date", y="Mean Monthly Amount")
```
```{r}
ggplot(data=filter(aggregated_transactions, industry == 6), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Industry 6 ONLY", x="Date", y="Mean Monthly Amount")
```
```{r}
ggplot(data=filter(aggregated_transactions, industry == 7 & location == 7), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Industry 7 and Location 7", x="Date", y="Mean Monthly Amount")
```
```{r}
ggplot(data=filter(aggregated_transactions, industry == 8 & location == 8), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Industry 8 and Location 8", x="Date", y="Mean Monthly Amount")
```
```{r}
ggplot(data=filter(aggregated_transactions, industry == 9 & location == 9), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Industry 9 and Location 9", x="Date", y="Mean Monthly Amount")
```
```{r}
ggplot(data=filter(aggregated_transactions, industry == 10 & location == 10), aes(x=date, y=monthly_amount, group=1)) + geom_line(color="red") + geom_point() + labs(title="Line plot of mean monthly amount for Industry 10 and Location 10", x="Date", y="Mean Monthly Amount")
```
##ALEX TO PLOT A GRAPH SHOWING SPLIT OF TEST AND TRAIN

### 2. a. iii) i) Trying to train the linear regression model with monthly_amount as the target

###ANT we make our first linear model WITH location 1 and industry 1 in it.  

```{r}
# Monthly_amount (or more specifically mean monthly amount) is the target variable whose values are to be modeled/predicted by other variables. Can't use location or industry as a predictor variable, so only other possible predictor is month number and/or date.

calculate_predictions <- function(df, industries, locations) {

  output = data.frame()
  
  #Want to make sure we capture a couple of months from the preceeding year to capture the seasonal data
  testingNum = 14
  #The test/train split should ensure training data has 'trainingMultiplier' times the number of entries of tesing entries
  trainingMultiplier = 3
  
  for (ind in industries) {
    for (loc in locations) {
      # create a subset of the data
      temp = df[df$industry == ind & df$location == loc, ]
      
      # Check to make sure you have at least 'trainingMultiplier' times the number of training rows than testing rows
      if (length(unique(temp$date)) >= trainingMultiplier*testingNum) {
        # train model
        
        #Arrange dataset by date
        arrange(temp, date)
        
        #Add a number to represent date order
        temp$time_number = c(1:nrow(temp))
        
        #Just to make sure, let's order by time_number
        arrange(temp, time_number)
        
        #Training number is the number of rows minus the testingNum
        trainingNum = nrow(temp) - testingNum
        
        #Training set is all rows from the start minus the number in the test set
        trainingSet = head(temp, trainingNum)
        #Testing set is the last 'testingNum' rows
        testingSet = tail(temp, testingNum)
        
        # check out the details
        model = lm(monthly_amount~month_number + time_number, data=trainingSet)
        
        # output a prediction
        temp$prediction = predict(model, temp)
        
        # calculate the mean standard error
        mse <- mean(residuals(model)^2)
        # calculate root mean squared error
        rmse <- sqrt(mse)
        
        # append error to the output data frame, including industry and location variables
        error = c(ind,loc,rmse)
      } else {
        # append entry to output data frame when not enough data to compute
        error = c(ind,loc,NA)
      }
      output = rbind(output, error)
    }
  }
  #Add column names to the output data frame
  colnames(output) <- c("industry","location", "rmse")
  
  return(output)
}

#Now, aggregate the data again tomake sure it's clean
aggregated_transactions <- aggregate_transactions("transactions.csv", T)

#Calculate the predictions for location 1 and industry 1
prediction <- calculate_predictions(aggregated_transactions, 4, 6)

prediction

```

###SO we then need to run these same linear models on the 
### Training the monthly_amount ~ month_number linear model

```{r}
# Run linear model with testing and training sets, with monthly_amount as target and month_number as the predictor
month_number.lm <- lm(monthly_amount~month_number, data=trainingSet)
  # plot the regression diagnostics.  Note the red line on two left plots is not straight
summary(month_number.lm)
```

## Assessing the monthly_amount ~ month_number linear model

The summary statistics let us see the size and significance of the fit for this model, which looks unlikely to be significant or a good fit. The low p-value of .09638 reflects this. The Adjusted R-Squared of 0.05211 further echoes this observation.

Residuals represent a lack of fit of a line to a model - which can be seen as the model's error - and you can see the minimum to maximum residuals range from -24168 to 17750.

```{r}
# let's try to see how linear the relationship between monthly amount and month number really is - or isn't
ggplot(data = trainingSet, mapping = aes(x = month_number, y = monthly_amount)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") + labs(title="The linear relationship between monthly amount and month number", x="Month Number", y="Monthly Amount")
```
## It's not the greatest fit, but there is a trend 
If we plot monthly_amount ~ month_number linear model using statistical diagnostic plots, we can get a better view of the sign, size and significance of the fit and further confirm that month number is not likely to be the best predictor of monthly sales amounts.

```{r}
# run diagnostic plots for the monthly_amount ~ month_number linear model
plot(month_number.lm)
```
### Training the monthly_amount ~ date linear model  

Now let's run a different linear model with monthly_amount as target and date as the predictor to see if it fits better.

```{r}
date.lm <- lm(monthly_amount~date, data=trainingSet)
  # plot the regression diagnostics to assess this fit versus previous model
summary(date.lm)
```
## Assessing the monthly_amount~date linear model 

```{r}
# let's try to see if there is a better linear relationship between monthly amount and date
ggplot(data = trainingSet, mapping = aes(x = date, y = monthly_amount)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") + labs(title="A stronger trend ", x="Date", y="Monthly Amount")
```

### Compared to monthly_amount ~ month_number linear model, this fits better
The residuals are closer together and Adjusted R-squared of 0.3446 is getting us closer but the monthly_amount ~ date linear model still isn't ideal.  

```{r}
#the diagnostic plots reveal this model to have more a more significant fit than the first linear model of monthly_amount ~ month_number.
plot(date.lm)
```

## Let's see if adding month number AND date work better in tandem
# Run linear model with monthly_amount as target and month number + date as the predictor

```{r}
month_number_and_date.lm <- lm(monthly_amount~month_number + date, data=trainingSet)
  # plot the regression diagnostics.   
summary(month_number_and_date.lm)
```
##This model of monthly_amount ~ month_number + date has not really improved  

The Adjusted R-squared of 0.3326 is not higher than our second model.

# Plot monthly_amount ~ month_number + date linear model
```{r}
plot(month_number_and_date.lm)
```

##2. a. iii) ii) How to split test and train sets
I chose to split the test and training data on years, and use the 2016 data as the test. I believed chronological split was the best way to try to predict the future monthly_amount. Unfortunately, the December seasonality is making the prediction hard.

The monthly_amount ~ date model was the best of the training data. But far from perfect.

##2. a. iv) Create a prediction for monthly_amount in December 2016
I really don't know how to work this out.

##2. b. Describe the model
Spit out the diagnostics for location 1, industry 1 BUT ALSO COMPUTE THE RMSE
## 2. b. i) How well does the model fit the data it is trained on in a statistical sense
RMSE should work for this - for regressions tasks, RMSE is a good choice. Could discuss RMSE vs MAE

Explain the feature selection and why you left most variables in

We will ultimately need three columns showing
Indsustry - Location - RMSE 
 
  
```{r}
## MAKE R CODE TO SPIT OUT INDUSTRY, LOCATION AND RMSE WITH A PREDICTION FOR DEC 
```
## 2. b. ii) How well does the model predict out of sample
Talk about RMSE here
```{r}
#TO DO
```
## 2. b. iii) Which features did you end up using in your final model?
Explain which variables went in and why. Justify your feature selection method using external resources.

## 3. Advanced model fitting
Describe what that means
## 3. a. Apply the modelling process you built for industry 1 and location 1 to all industries and locations programmatically

```{r}
#TO DO
```
## 3. a. Apply the modelling process you built for industry 1 and location 1 to all industries and locations programmatically
Describe what that means
```{r}
#Monthly_amount (or more specifically mean monthly amount) is the target variable whose values are to be modeled/predicted by other variables. Can't use location or industry as a predictor variable, so only possible predictors is month number and/or date.

#Now, aggregate the data again tomake sure it's clean
aggregated_transactions <- aggregate_transactions("transactions.csv", T)

industries <- sort(unique(aggregated_transactions$industry))
locations <- sort(unique(aggregated_transactions$location))

#Calculate the predictions for location 1 and industry 1
all_predictions <- calculate_predictions(aggregated_transactions, industries, locations)

all_predictions <- arrange(all_predictions, rmse)

all_predictions

```
## 3. b. Calculate your evaluation measure for the training data and your testing data for all models
This is where i talk more about RMSE
## 3. b. Identify the two worst industries and two locations for which your method performs worst.
Describe what that means
## 3.c. What is causing the poor performance of the two indsutries and two locations that do badly? How might you fix it in the future?
Describe what that means
## 4. Reporting - see separate Word document with report. Include an appendix and references in this main report.
 
