---
title: "DAM At1 EDA & Data Cleaning"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#note that correct working directory is setwd("/Users/alex/Projects/DAMAT1")
#load libraries 
#library(tidyverse)
#library(modelr)
library(readr)
#library(ggplot2)
#library(dplyr)
#library(ggExtra)
#library(dygraphs)
```

### AT1 Spring 2018: Part A
## Data Cleaning and Exploratory Data Analysis on "transactions.csv" dataset
I will explore the data using predictor variables like date, location, industry and customer_id.
```{r}

#import dataset using read.csv
transactions <- read_csv("transactions.csv")

```

# 1. a. Cleaning the data

It was noted that readr did not identify the date column as a data type of date. In order to give the date colum a correct datatype of date, the parse_date function is used to parse the values in the date colum as dates.

To ensure that any other artefacts are cleaned up, a new 'clean' csv is written and read back in (overwriting the original transaction variable) and the column types are explicitly set.

```{r}

# Use the date parser to get the date data into an understandable format
transactions$date <- parse_date(transactions$date, format="%d/%m/%y")

# Write a new 'clean' csv with correct date formats and remove any artefacts from opening in Excel
write_csv(transactions, "transactions_clean.csv")

# Read in the clean data, overwriting transactions and explicitly declaring the data types for each column
transactions <- read_csv("transactions_clean.csv",
                         col_types = cols(
                           date = col_date(format = ""),
                           customer_id = col_character(),
                           industry = col_integer(),
                           location = col_integer(),
                           monthly_amount = col_double()
                           )
                         )
# Verify transactions data is using the correct data types
spec(transactions)

# View the data structure
View(transactions)
```

#1. b. Description of 2 insights just from EDA

##show histograms of industry and location variables
```{r}
par(mfrow=c(1,2))
hist(transactions$industry, main = "Histogram of number of transactions by Industry", xlab="Industry", ylab="Number of transactions", xlim = c(0,10), ylim=c(0,50000), las=0)
hist(transactions$location, main = "Histogram of number of transactions by Location", xlab="Location", ylab="Number of transactions", xlim = c(0,10), ylim=c(0,50000), las=0)
```
# More visual exploration
It looks like locations one and two account for the greatest majority of monthly transactions and there's something about industry code 6 that's very distinct from the other industries.

```{r}
# Let's mutate the dataset to create a new column called 'monthly_amount_above_mean'
transactions <- transactions %>%
  # creating a new variable to classify transaction column
  mutate(monthly_amount_above_mean = ifelse(monthly_amount > mean(monthly_amount), TRUE, FALSE))
```

```{r}
# plot a bar chart by monthly amounts by date
qplot(x = date, fill = monthly_amount_above_mean, data = transactions, geom = "bar", main = "Number of transactions by month, showing proportion above the mean")
```

This reveals most monthly amounts are below the mean and the monthly amounts grow steadily rather than steeply. Let's now look at it by industry and location, too
```{r}
# plot a bar chart by monthly amounts by location
qplot(x = location, fill = monthly_amount_above_mean, data = transactions, geom = "bar", main = "NUmber of transactions by location, showing proportion above the mean")

```
#So what about by industry?

```{r}
# plot a bar chart by monthly amounts by industry
qplot(x = industry, fill = monthly_amount_above_mean, data = transactions, geom = "bar", main = "Number of transactions by industry, showing proportion above the mean")
```
##Some insights
Location 1 and 2 have the highest volumes of monthly_amounts, as well as the most above-average monthly amounts.

Industry 6 has the least amount of sales, but when they have them, they are all above the mean of all sales amounts.

Industries 3, 5 and 9 look to have a greater proportion of their transactions above the mean average monthly amount

Let's aggregate the data and see if we can get more insights
```{r}
# Aggregate the data, grouping by date, industry and location, and calculating the mean monthly_amount
aggregated_transactions <- transactions %>%
  group_by(date, industry, location) %>%
  summarize(mean_monthly_amount = mean(monthly_amount, na.rm = TRUE))

# View the data
aggregated_transactions

```


```{r}
#let's plot the location variable against the average monthly amount , ylim=c(0,1000000)
#industry_boxplot <- boxplot(mean_monthly_amount ~ industry, data = aggregated_transactions) +
#      geom_point()
```

```{r}
#let's try to make a time series plot
#library(ggfortify)
#autoplot(~ monthly_amount) + labs(title="Monthly Sales Amount")
```


```{r}
#let's visualise some of the first import of the data as a scatterplot
ggplot(transactions, aes(x=industry, y=monthly_amount, color=location)) + 
  geom_point() + geom_smooth()
```
fG
GA

```{r}
#let's visualise the second data import transactions as a scatterplot to see if there's a difference
ggplot(transactions, aes(x=industry, y=monthly_amount, color=location)) + 
  geom_point() + geom_smooth()
```
No difference! These plots are identical, even though we are using two slightly different column structures of the same data.

## More exploration using GGPLOT2 & Dp

That's a slightly ridiculous plot, but it shows there is something going up with the industry number 6:
```{r}
##Take the column "industry" and make a histogram of it
qplot(transactions$industry, geom="histogram", binwidth=0.5) 
```
The histogram shows the industry variable is front weighted, I THINK to a poisson distribution. Still not super helpful
```{r}
##Take a different column "location" and make a histogram of it
qplot(transactions$location, geom="histogram", binwidth=0.5) 
```
Location data is also front weighted.

```{r}
#get the sum of all monthly amounts by location
barplot <- ggplot(transactions, aes(x=monthly_amount, y=location)) + geom_bar(stat = "identity")
```



```{r}
#another scatterplot
ggplot(transactions) + geom_point(aes(x=location, y=monthly_amount, color=industry)) + geom_smooth(aes(x=location, y=monthly_amount, color=industry)) # Specifying the aesthetics inside the geoms.
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
#2. a. ii. Create a line plot of the variable monthly_amount, for industry = 1 and location = 1

ggplot(data=filter(aggregated_transactions, industry == 1 & location == 1), aes(x=date, y=mean_monthly_amount, group=1)) + geom_line(color="red") + geom_point()
```



